---
title: "STAT8810 HW3-Question4"
author: "Hengrui Luo"
date: "November 8, 2017"
output:
  word_document: default
  pdf_document: default
  html_document: default
---
```{r,eval=F}
##########################################################################################################################
# Question 4
##########################################################################################################################
# For STAT8810 HW 3
# Fall, 2017
# M.T. Pratola, modified by Hengrui Luo
library(fOptions)
library(mvtnorm)
source("dace.sim.R")
source("regress.original.R")
load("co2holdout.dat_.txt")
#install.packages("tgp")
library(tgp)
# try a treed GP LLM
# best to use the "b0" beta linear prior to capture common
# common linear process throughout all regions (using the
# ellipses "...") 
co2df<-as.data.frame(co2holdout)[,]
cat("\nReading response\n")
n=dim(co2df)[1]
k=2
p=1
simtrue=list(rhotrue=0,lambdaytrue=0,betatrue=rep(0,p))
pi=list(az=5,bz=5,rhoa=rep(1,k),rhob=rep(5,k),Fz=matrix(1,nrow=n,ncol=p),mu=rep(1,p),lambdamu=1 )
design=as.matrix(co2df[,1:2])
l1=list(m1=outer(design[,1],design[,1],"-"))
l2=list(m2=outer(design[,2],design[,2],"-"))
l.dez=list(l1=l1,l2=l2)

# now set up our observed data:
y=co2df[,3]
```

```{r,eval=F}
##########################################################################################################################
# (a) Fit the CO2 data with BGP in Question 3
##########################################################################################################################
#source("regression.r")

# Run MCMC using a proposal width of 1e-5 for the rho parameters.
# mh=list(rr=0.01)
# fit=regress(y,l.dez,5000,pi,mh,last=2000,adapt=FALSE)
# plotfit2(fit)
#Why bother non-adaptive when you have such an option...

mh=list(rr=0.01)
fit2=regress(y,l.dez,5000,pi,mh,last=2000,adapt=TRUE)
plotfit2(fit2)
# plotfit2 plot trace plot for each 
# Predict the last fit on a grid of 100 equally spaced points
grid.num=11
grid=as.matrix(seq(0,1,length=grid.num))
grid=expand.grid(grid,grid)
#Create a 2 dim grid matrix.
grid=as.matrix(grid)
#two dim grid at each design location and calculate their relative distance norm.
design.all=rbind(design,grid)
l1=list(m1=outer(design.all[,1],design.all[,1],"-"))
l2=list(m2=outer(design.all[,2],design.all[,2],"-"))
l.v=list(l1=l1,l2=l2)

fitp=predict(l.v,fit2,eps.yy=1e-4,force=F)

library(rgl)
# Plot the posterior predictions using mean and sd to quantify uncertainty (pointwise), we can use 97.5% and 2.5% posterior quantiles as cutpoints of credible intervals.
ysd=apply(fitp$preds,2,sd)
yq1=apply(fitp$preds,2,quantile,probs=.025)
yq2=apply(fitp$preds,2,quantile,probs=.975)
ym=apply(fitp$preds,2,mean)

# Plot fit and uncertainties
persp3d(x=seq(0,1,length=grid.num),y=seq(0,1,length=grid.num),z=matrix(ym,grid.num,grid.num),
        col="grey",xlab="stack_inerts",ylab="time",zlab="CO2")
persp3d(x=seq(0,1,length=grid.num),y=seq(0,1,length=grid.num),
        z=matrix(yq1,grid.num,grid.num),col="green",add=TRUE,alpha=.5)
persp3d(x=seq(0,1,length=grid.num),y=seq(0,1,length=grid.num),
        z=matrix(yq2,grid.num,grid.num),col="green",add=TRUE,alpha=.5)
plot3d(x=co2df$stack_inerts,y=co2df$time,z=co2df$co2,add=TRUE,size=1,type='s',col='red')

persp3d(x=seq(0,1,length=grid.num),y=seq(0,1,length=grid.num),z=matrix(ym,grid.num,grid.num),
        col="grey",xlab="stack_inerts",ylab="time",zlab="CO2")
persp3d(x=seq(0,1,length=grid.num),y=seq(0,1,length=grid.num),
        z=matrix(ym-ysd,grid.num,grid.num),col="blue",add=TRUE,alpha=.5)
persp3d(x=seq(0,1,length=grid.num),y=seq(0,1,length=grid.num),
        z=matrix(ym+ysd,grid.num,grid.num),col="blue",add=TRUE,alpha=.5)
plot3d(x=co2df$stack_inerts,y=co2df$time,z=co2df$co2,add=TRUE,size=1,type='s',col='red')

rgl.viewpoint(30,30,60)
rgl.snapshot(filename = "BGP1.png")
rgl.viewpoint(45,-90,60)
rgl.snapshot(filename = "BGP2.png")
rgl.viewpoint(90,-90,60)
rgl.snapshot(filename = "BGP3.png")
#Here we completely plot the uncertainty quantification 
```
(We omit the output here since it is nothing much different from output from Question 3)

```{r,eval=F}
##########################################################################################################################
# (b) Fit the CO2 data with BTGP (Bayesian Treed GP)
##########################################################################################################################
library(tgp)
grid.num=11
grid=as.matrix(seq(0,1,length=grid.num))
grid=expand.grid(grid,grid)
#Create a 2 dim grid matrix.
grid=as.matrix(grid)
X <- co2df[,1:2]
XX <- grid
Z <- co2df[,3]
fit.tgp <- btgp(X=X, Z=Z, XX=XX,trace=T) 
plot(fit.tgp) # plot the surface
tgp.trees(fit.tgp) # plot the MAP trees
#Diagnostics
par(mfrow=c(1,2))
plot(fit.tgp$trace$hier$s2.a0,type='l')
plot(fit.tgp$trace$preds$Zp.ks2$XX1,type='l')
# Plot fit

ysd=fit.tgp$ZZ.vark
yq1=fit.tgp$ZZ.q1
yq2=fit.tgp$ZZ.q2
ym =fit.tgp$ZZ.mean
#Here we can also use 5% and 95% quantile instead of variance as uncertainty quantification interval, as suggested by TGP. What we get directly from btgp function is the 2.5% and 97.5% quantile of the posterior samples.
persp3d(x=seq(0,1,length=grid.num),y=seq(0,1,length=grid.num),z=matrix(ym,grid.num,grid.num),
        col="grey",xlab="stack_inerts",ylab="time",zlab="CO2")
plot3d(co2df,add=TRUE)

# Plot fit and uncertainties
persp3d(x=seq(0,1,length=grid.num),y=seq(0,1,length=grid.num),z=matrix(ym,grid.num,grid.num),
        col="grey",xlab="stack_inerts",ylab="time",zlab="CO2")
persp3d(x=seq(0,1,length=grid.num),y=seq(0,1,length=grid.num),
        z=matrix(yq1,grid.num,grid.num),col="green",add=TRUE,alpha=.5)
persp3d(x=seq(0,1,length=grid.num),y=seq(0,1,length=grid.num),
        z=matrix(yq2,grid.num,grid.num),col="green",add=TRUE,alpha=.5)
plot3d(x=co2df$stack_inerts,y=co2df$time,z=co2df$co2,add=TRUE,size=1,type='s',col='red')

persp3d(x=seq(0,1,length=grid.num),y=seq(0,1,length=grid.num),z=matrix(ym,grid.num,grid.num),
        col="grey",xlab="stack_inerts",ylab="time",zlab="CO2")
persp3d(x=seq(0,1,length=grid.num),y=seq(0,1,length=grid.num),
        z=matrix(ym-ysd,grid.num,grid.num),col="blue",add=TRUE,alpha=.5)
persp3d(x=seq(0,1,length=grid.num),y=seq(0,1,length=grid.num),
        z=matrix(ym+ysd,grid.num,grid.num),col="blue",add=TRUE,alpha=.5)
plot3d(x=co2df$stack_inerts,y=co2df$time,z=co2df$co2,add=TRUE,size=1,type='s',col='red')

rgl.viewpoint(30,30,60)
rgl.snapshot(filename = "TGP1.png")
rgl.viewpoint(45,-90,60)
rgl.snapshot(filename = "TGP2.png")
rgl.viewpoint(90,-90,60)
rgl.snapshot(filename = "TGP3.png")
```
burn in:  
**GROW** @depth 0: [1,0.5], n=(208,164)  
**PRUNE** @depth 0: [1,0.5]  
r=1000 d=[0.00259203 0.220003]; n=372  
**GROW** @depth 0: [2,0.2], n=(87,285)  
r=2000 d=[0.104118 0.458754] [0.00263412 0.936753]; n=(87,285)  
  
Sampling @ nn=121 pred locs:  
r=1000 d=[0.614422 0.0753061] [0.00240447 1.06558]; mh=2 n=(87,285)  
**GROW** @depth 1: [1,0.498798], n=(46,41)  
r=2000 d=[0.0699097 0.036066] [0.697349 0.0354746] [0.0025254 0.729628]; mh=3 n=(46,41,285)  
r=3000 d=[0.197858 0.0750243] [0.789468 0.125436] [0.00291456 0.935195]; mh=3 n=(46,41,285)  
r=4000 d=[0.00170656 0.0173697] [1.0251 0.0667556] [0.00289671 1.28135]; mh=3 n=(56,49,267)  
r=5000 d=[0.00118787 0.0603981] [0.815149 0.0383773] [0.00294744 0.895394]; mh=3 n=(46,41,285)  
Grow: 0.8746%, Prune: 0.3831%, Change: 20.7%, Swap: 0%  

```{r,eval=F}
##########################################################################################################################
# (c) Fit the CO2 data with BayesTree (Bayes single tree, m=1)
##########################################################################################################################
# Fit the BART model with multiple m trees.
y=co2df$co2
x=co2df[,1:2]
preds=as.data.frame(expand.grid(seq(0,1,length=grid.num),
                                seq(0,1,length=grid.num)))
colnames(preds)=colnames(x)
shat=sd(y)
# Try m=1 tree, single tree GP
m=1
# And k=2
k=2
# And nu=1, q=.99
nu=3
q=0.99
# And numcuts=1000
nc=1000
# MCMC settings
N=1000
burn=1000
fit.BART=bart(x.train=x,y.train=y,x.test=preds,sigest=shat,sigdf=nu,sigquant=q,
              k=k,power=beta,base=alpha,ntree=m,numcut=nc,
              ndpost=N,nskip=burn)

# Plot posterior samples of sigma
plot(fit.BART$sigma,type='l',xlab="Iteration",
     ylab=expression(sigma))

# Plot fit
ym =fit.BART$yhat.test.mean
ysd=apply(fit.BART$yhat.test,2,sd)
persp3d(x=seq(0,1,length=grid.num),y=seq(0,1,length=grid.num),z=matrix(ym,grid.num,grid.num),
        col="grey",xlab="stack_inerts",ylab="time",zlab="CO2")
plot3d(co2df,add=TRUE)
#Here we can either choose 2.5%,97.5% posterior sample quantiles as uncertainty quantification or +/- 2sd, but we choose 95% credible interval according to the HW requirement.
# Plot fit and uncertainties
persp3d(x=seq(0,1,length=grid.num),y=seq(0,1,length=grid.num),z=matrix(ym,grid.num,grid.num),
        col="grey",xlab="stack_inerts",ylab="time",zlab="CO2")
persp3d(x=seq(0,1,length=grid.num),y=seq(0,1,length=grid.num),
        z=matrix(yq1,grid.num,grid.num),col="green",add=TRUE,alpha=.5)
persp3d(x=seq(0,1,length=grid.num),y=seq(0,1,length=grid.num),
        z=matrix(yq2,grid.num,grid.num),col="green",add=TRUE,alpha=.5)
plot3d(x=co2df$stack_inerts,y=co2df$time,z=co2df$co2,add=TRUE,size=1,type='s',col='red')

persp3d(x=seq(0,1,length=grid.num),y=seq(0,1,length=grid.num),z=matrix(ym,grid.num,grid.num),
        col="grey",xlab="stack_inerts",ylab="time",zlab="CO2")
persp3d(x=seq(0,1,length=grid.num),y=seq(0,1,length=grid.num),
        z=matrix(ym-ysd,grid.num,grid.num),col="blue",add=TRUE,alpha=.5)
persp3d(x=seq(0,1,length=grid.num),y=seq(0,1,length=grid.num),
        z=matrix(ym+ysd,grid.num,grid.num),col="blue",add=TRUE,alpha=.5)
plot3d(x=co2df$stack_inerts,y=co2df$time,z=co2df$co2,add=TRUE,size=1,type='s',col='red')
plot(fit.BART)

rgl.viewpoint(30,30,60)
rgl.snapshot(filename = "BART1.png")
rgl.viewpoint(45,-90,60)
rgl.snapshot(filename = "BART2.png")
rgl.viewpoint(90,-90,60)
rgl.snapshot(filename = "BART3.png")
```
Running BART with numeric y  
  
number of trees: 1  
Prior:  
	k: 2.000000  
	degrees of freedom in sigma prior: 3  
	quantile in sigma prior: 0.990000  
	power and base for tree prior: 2.000000 0.950000  
	use quantiles for rule cut points: 0  
data:  
	number of training observations: 372  
	number of test observations: 121  
	number of explanatory variables: 2  
  

Cutoff rules c in x<=c vs x>c  
Number of cutoffs: (var: number of possible c):  
(1: 1000) (2: 1000)  


Running mcmc loop:  
iteration: 100 (of 2000)  
iteration: 200 (of 2000)  
iteration: 300 (of 2000)  
iteration: 400 (of 2000)  
iteration: 500 (of 2000)  
iteration: 600 (of 2000)  
iteration: 700 (of 2000)  
iteration: 800 (of 2000)  
iteration: 900 (of 2000)  
iteration: 1000 (of 2000)  
iteration: 1100 (of 2000)  
iteration: 1200 (of 2000)  
iteration: 1300 (of 2000)  
iteration: 1400 (of 2000)  
iteration: 1500 (of 2000)  
iteration: 1600 (of 2000)  
iteration: 1700 (of 2000)  
iteration: 1800 (of 2000)  
iteration: 1900 (of 2000)  
iteration: 2000 (of 2000)  
time for loop: 0  

Tree sizes, last iteration:  
25 Variable Usage, last iteration (var:count):  
(1: 11) (2: 13)   
DONE BART 11-2-2014  


```{r,eval=F}
##########################################################################################################################
# (d) Fit the CO2 data with BART (Bayesian Additive Regression Tree)
##########################################################################################################################
# Fit the BART model with multiple m trees.
y=co2df$co2
x=co2df[,1:2]
preds=as.data.frame(expand.grid(seq(0,1,length=grid.num),
                                seq(0,1,length=grid.num)))
colnames(preds)=colnames(x)
shat=sd(y)
# Try m=200 trees, the recommended default
m=200
# And k=2
k=2
# And nu=1, q=.99
nu=3
q=0.99
# And numcuts=1000
nc=1000
# MCMC settings
N=1000
burn=1000
fit.BARTm=bart(x.train=x,y.train=y,x.test=preds,sigest=shat,sigdf=nu,sigquant=q,
         k=k,power=beta,base=alpha,ntree=m,numcut=nc,
         ndpost=N,nskip=burn)

# Plot posterior samples of sigma
plot(fit.BARTm$sigma,type='l',xlab="Iteration",
     ylab=expression(sigma))

# Plot fit
yq1=apply(fit.BARTm$yhat.test,2,quantile,probs=.025)
yq2=apply(fit.BARTm$yhat.test,2,quantile,probs=.975)
ysd=apply(fit.BARTm$yhat.test,2,sd)
ym =fit.BARTm$yhat.test.mean
#Here we can either choose 2.5%,97.5% posterior sample quantiles as uncertainty quantification or +/- 2sd, but we choose 95% credible interval according to the HW requirement.
persp3d(x=seq(0,1,length=grid.num),y=seq(0,1,length=grid.num),z=matrix(ym,grid.num,grid.num),
        col="grey",xlab="stack_inerts",ylab="time",zlab="CO2")
plot3d(co2df,add=TRUE)

# Plot fit and uncertainties
persp3d(x=seq(0,1,length=grid.num),y=seq(0,1,length=grid.num),z=matrix(ym,grid.num,grid.num),
        col="grey",xlab="stack_inerts",ylab="time",zlab="CO2")
persp3d(x=seq(0,1,length=grid.num),y=seq(0,1,length=grid.num),
        z=matrix(yq1,grid.num,grid.num),col="green",add=TRUE,alpha=.5)
persp3d(x=seq(0,1,length=grid.num),y=seq(0,1,length=grid.num),
        z=matrix(yq2,grid.num,grid.num),col="green",add=TRUE,alpha=.5)

persp3d(x=seq(0,1,length=grid.num),y=seq(0,1,length=grid.num),z=matrix(ym,grid.num,grid.num),
        col="grey",xlab="stack_inerts",ylab="time",zlab="CO2")
persp3d(x=seq(0,1,length=grid.num),y=seq(0,1,length=grid.num),
        z=matrix(ym-ysd,grid.num,grid.num),col="blue",add=TRUE,alpha=.5)
persp3d(x=seq(0,1,length=grid.num),y=seq(0,1,length=grid.num),
        z=matrix(ym+ysd,grid.num,grid.num),col="blue",add=TRUE,alpha=.5)
plot3d(x=co2df$stack_inerts,y=co2df$time,z=co2df$co2,add=TRUE,size=1,type='s',col='red')
plot(fit.BART)

rgl.viewpoint(30,30,60)
rgl.snapshot(filename = "BARTm1.png")
rgl.viewpoint(45,-90,60)
rgl.snapshot(filename = "BARTm2.png")
rgl.viewpoint(90,-90,60)
rgl.snapshot(filename = "BARTm3.png")
```
Running BART with numeric y  

number of trees: 200  
Prior:  
	k: 2.000000  
	degrees of freedom in sigma prior: 3  
	quantile in sigma prior: 0.990000  
	power and base for tree prior: 2.000000 0.950000  
	use quantiles for rule cut points: 0  
data:  
	number of training observations: 372  
	number of test observations: 121  
	number of explanatory variables: 2  


Cutoff rules c in x<=c vs x>c  
Number of cutoffs: (var: number of possible c):  
(1: 1000) (2: 1000)   


Running mcmc loop:  
iteration: 100 (of 2000)  
iteration: 200 (of 2000)  
iteration: 300 (of 2000)  
iteration: 400 (of 2000)  
iteration: 500 (of 2000)  
iteration: 600 (of 2000)  
iteration: 700 (of 2000)  
iteration: 800 (of 2000)  
iteration: 900 (of 2000)  
iteration: 1000 (of 2000)  
iteration: 1100 (of 2000)  
iteration: 1200 (of 2000)  
iteration: 1300 (of 2000)  
iteration: 1400 (of 2000)  
iteration: 1500 (of 2000)  
iteration: 1600 (of 2000)  
iteration: 1700 (of 2000)  
iteration: 1800 (of 2000)  
iteration: 1900 (of 2000)  
iteration: 2000 (of 2000)  
time for loop: 69  
  
Tree sizes, last iteration:  
2 1 2 3 2 2 3 2 3 2 3 3 2 2 3 3 3 3 3 2   
2 2 2 2 2 2 2 2 3 3 3 2 2 2 2 2 3 2 3 2   
5 4 2 2 1 2 3 2 2 3 3 2 2 3 4 2 3 2 4 3   
3 2 3 3 2 2 2 2 2 2 3 3 1 2 2 2 3 3 2 2   
3 2 2 2 2 2 2 2 2 3 3 3 2 2 2 2 2 2 2 2   
3 2 3 3 2 2 3 2 2 3 2 2 2 2 2 3 3 2 2 2   
2 2 2 2 3 2 2 2 2 1 2 3 2 3 1 2 2 2 1 2   
2 2 4 2 2 2 3 2 1 3 4 3 2 2 4 4 2 3 3 4   
2 1 2 2 2 3 4 2 2 2 2 3 2 2 2 3 2 3 3 3   
4 2 2 3 2 3 2 4 2 3 2 3 2 2 2 2 2 2 2 2   
Variable Usage, last iteration (var:count):  
(1: 138) (2: 136)   
DONE BART 11-2-2014  
